Make any stupid plot you want

### 2021-04-29 draft/sections/intro.tex
tools

\begin{figure}[H]
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/table.png}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/graph.png}
    \end{subfigure}
    \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables, images, or networks. }
    \label{fig:intro:data}
\end{figure}

One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures. These libraries can often assume the expected data structure because they are domain specific, and that is the common data structure in that domain.  For users who generally work in one domain, such as the data, networks, or graphs shown in \autoref{fig:intro:data}, this well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a very coherent user experience\cite{heerDeclarative2010}. But, for developers who want to build new visualizations on top of these libraries, they must work around the existing assumptions, sometimes in ways that break the model the libraries are developed around.

For example, many domain specific libraries integrate computation into the visualization, for example libraries based that assume all data is a relational database. This assumption is core to tools influenced by APT, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}, such as ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. Since these libraries represent all data as a table, and computations on tables are fairly well defined\cite{ullmanFirstCourseDatabase2008}, they can include computations on the table with a fair bit of confidence that the computation is accurate. Since most computations are specific to domains, general purpose block libraries can not make this assumption; instead a goal of this model is to identify which computations are specifically part of the visual encoding - for example mapping data to a color-and which are manipulations on the data. Disentangling the computation from the visual transforms allows us to determine whether the visualization library needs to handle them or if they can be more efficiently computed by the data container.

A different class of user facing tools are those that support images, such as ImageJ\cite{schneiderNIHImageImageJ2012} or Napari\cite{nicholas_sofroniew_2021_4533308}. These tools often have some support for visualizing non image components of a complex data set, but mostly in service to the image being visualized. These tools are ill suited for general purpose libraries that need to support data other than images because the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure. Even the digital humanities oriented ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}, which supports some non-image aggregate reporting charts, is still built around image data as the primary input. The need to visualize and manipulate graphs has spawned tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008}. As with tables and images, extending network libraries to work with other types of data either require breaking their internal model of how data is structured and what transformations of the data are allowable or growing a model for other types of data structures alongside the network model.

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
    \caption{One way to describe data is by the connectivity of the points in the dataset. A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid. This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
    \label{fig:intro:data:format}
\end{figure}
Many building block libraries carry multiple models of data internally because they cannot assume a data structure. Algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}, and by definition building block libraries try to  provide the components to build any sort of visualization. Matplotlib, D3\cite{bostockDataDrivenDocuments2011}, and VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. Where GoG and ImageJ type libraries have coherant APIs for their visualization tools because the data structure is the same, the APIs for visualizations in Matplotlib, D3, and VTK are significantly dependent on the structure of the data it expects. VTK has codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}. This API choice can lead to visualizations that break \textit{continuity} when fed into visualizations with different assumptions about structure. The lack of consistent data model can also mean no consistent way of updating the data and therefore no way of guaranteeing that the views are in sync, in visualizations that consistent of multiple views of the same datasource, such as dashboards\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007}. To resolve this issue, our functional model takes as input a structure aware data abstraction general enough to provide a common interface for many different types of visualization.

>>>>

\begin{figure}[H]
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/table.png}
        \caption{}
        \label{fig:intro:table}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
        \caption{}
        \label{fig:intro:image}
    \end{subfigure}
    \begin{subfigure}{.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/graph.png}
        \caption{}
        \label{fig:intro:graph}
    \end{subfigure}
    \caption{Visualization libraries, especially ones tied to specific domains, tend to be architectured around a core data structure, such as tables \autoref{fig:intro:table}, images \autoref{fig:intro:image}, or networks \autoref{fig:intro:graph}.
    %%revise as combined fig?
    }
    \label{fig:intro:data}
\end{figure}

Most information visualization software design patterns are tuned to specific data structures and domains, as categorized by Heer and Agrawala\cite{HeerSoftware2006}. For users who generally work in one domain--such relational tables (\autoref{fig:intro:table}), images (\autoref{fig:intro:image}), or graphs (\autoref{fig:intro:graph})--well defined data space (and corresponding visual space\cite{chiTaxonomyVisualizationTechniques2000}) often yields a coherent user experience\cite{heerDeclarative2010}. This coherant experience is often not extensible; developers who want to build new visualizations on top of these libraries must work around the existing assumptions, sometimes in ways that break the architecture model of the libraries. For example tools influenced by APT that assume that data is a relational table integrate computation into the visualization pipeline. This is a wide array of tools, including Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007} and the Grammar of Graphics\cite{wilkinsonGrammarGraphics2005} inspired ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. Since these libraries represent data as a table, and computations on tables are well defined\cite{ullmanFirstCourseDatabase2008}, these libraries implement computation as part of the visualization process. This is also true of tools that support images, such as the biology oriented ImageJ\cite{schneiderNIHImageImageJ2012} and  Napari\cite{nicholas_sofroniew_2021_4533308} or the digital humanities ImageJ macro ImagePlot\cite{studiesCulturevisImageplot2021}. While these tools often have some support for visualizing non image components of the data, the architecture is oriented towards building plugins into the existing system \cite{WritingPlugins} where the image is the core data structure. Tools like Gephi\cite{bastianGephiOpenSource2009}, Graphviz\cite{ellsonGraphvizOpenSource2002}, and Networkx\cite{HagbergExploringNetwork2008} are used to  visualize and manipulate graphs. As with tables and images, extending network libraries to work with other types of data either requires breaking the libraries internal model of how data is structured and what data transformations are allowed or implementing a model for other types of data structures alongside the network model. Our model aims to identify which computations are manipulations of the data and which are necessary for the visual encoding; for example data is often aggregated for a bar chart but it is not required, while a boxplot by definition requires computing distrobution statistics\cite{wickham40YearsBoxplots2011}. Disentangling the computation from the visual transforms allows us to determine whether the visualization library should implement computations or if they should be implemented in data space.

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
    \caption{The \textit{continuity} of the data is often used to classify data structures. For example, a relational database often consists of discrete unconnected points, while an image is an implicitly connected 2D grid. This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
    \label{fig:intro:data:format}
\end{figure}

Visualizations assume the structure of the input data, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighlevel2004}, which leads to many building block libraries implementing multiple models of data to support these different visualizations. For example, in Matplotlib, D3\cite{bostockDataDrivenDocuments2011} VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} and MayaVi\cite{RamachandranMayaVI2011}, every plot is defined in terms of the \textit{continuity} of the data it expects as input. VTK has explicitly codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}. Downstream library developers impose some coherency by writing domain specific libraries with assumed data structures on top of the building block libraries--for example Seaborn\cite{waskom2020seaborn} and Titan\cite{brianwylieUnifiedToolkitInformation2009} assume a relational database, xarray\cite{hoyer2017xarray} and ParaView\cite{ahrens2005paraview} assume a data cube--but must work around the incoherencies in the building block libraries to do so. Our model navigates the tradeoff between coherency and extensibility by proposing functional composable well constrained visual components that take as input a structure aware data abstraction general enough to provide a common interface for many different types of visualizations.
